Computing likelihoods under the SMC for a general class of ARGs.

### 2024-03-22 paper.tex

The contributions to the likelihood from coalescent and recombination events
is straightforward, since each possible coalescence is associated with rate $\lambda$,
and each recombination with rate $r$.
Under the SMC, we only ever have two lineages coalescing at a time;
however, inferred ARGs may have more-than-binary mergers.
We treat these as if there were zero-length edges
between a sequence of mergers;
the order of these unresolved mergers does not affect the likelihood.
(Under the SMC the likelihood of a zero-length edge is zero;
but this is not special: since edge lengths are continuously distributed,
the likelihood of any particular value is also zero.)
Let $\deg_o(n) = \#\{e_i \in E : p_i=n\}$ be the out-degree of a node $n$,
i.e., the number of edges having $n$ as a parent.
Since mergers are binary in the SMC,
this must have been the result of $\deg(n) - 1$ coalescences,
and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood.
Similarly,
let $\deg_i(n) = \#\{e_i \in E : c_i=n\}$ be the in-degree of a node $n$,
i.e., the number of edges having $n$ as a child.
Each additional edge that is ancestral to a given node
implies one additional ancestral recombination event,
and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood.
Put together, these two contribute
\begin{equation}\label{eq:coal}
    \prod_{n \in N} \lambda^{\deg_o(n)-1} r^{\deg_i(n)-1}  
    =
    (\lambda r)^{|E|-|N|} 
\end{equation}
to the likelihood.
(The simple form follows because each edge contributes exactly one
to some node's out-degree and some node's in-degree.)

@@

The likelihood of ARG defined by the set of edges $E$ and nodes $N$
given parameters $\theta$ then is
\begin{equation}\label{eq:full-lik}
    \mathcal{L}(\mathcal{G}|\theta)
    =
    (r\lambda)^{|E|-|N|} \prod_{(c, p) \in E} A_e(\theta) B_e(\theta)
\end{equation}

@@

Although the argument outlined in the previous paragraph is valid for any ARG, here we
will describe how to (efficiently) extract the information required to associate a
tree sequence encoding of an ARG with a likelihood under the SMC (see fig. \ref{fig:smc-unary}).
The tree sequence format allows us to sequentially recover the local trees
along the genome very efficiently, and in a way that allows us to reason about
the differences between those trees \citep{kelleher_efficient_2016, ralph_efficiently_2020}.
The order of the required edge deletions and insertions to move from one local tree $T_1$ to
the next, $T_2$ imposes the required strict total ordering
(see definition \ref{def:coal}) on all
lineages in the graph needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
Indeed, following the removal of edges that do not persist in $T_2$,
we can postulate that the lineage associated with each subsequently added edge
can only coalesce with the lineages represented by the subset of the
visited edges that make up the (partially) reconstructed local tree $T_2$.

This ensures we can associate a likelihood
with all tree sequences in a single pass over all edges.
More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
the exact number of extant lineages in each internode interval (see fig \ref{fig:algo}).
When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along the internode intervals
spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
For each subsequently inserted edge, we first compute the likelihood by computing the integral
in eq \ref{eq:full-lik} using $I$ and then update the counts. We further keep track of
the last parent a child node was connected to to detect recombination events.

Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
through time by intersecting the internode intervals with the rate change intervals.
Rate changes along the genome can equally be incorporated (see \ref{sec:slicing}).

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG in both
time as well as along the genome. Our approach therefore strengthens
the inherent ability of ARGs to provide us with temporal
resolution on the inferred genealogical history of a sample.
It also opens up the opportunity for more efficient optimization
strategies during inference given that each slice of the ARG can be
processed independently.

From a data structure perspective, we can compute a likelihood
under the SMC for any valid tree sequence.
The \tskit library provides a whole range
of subsetting operations on tree sequences allowing us to slice
the ARG in time and/or space, as well as reduce the number of
tracked samples. Any such subset of the tree sequence is itself
a valid tree sequence that can again be used to compute a
likelihood for. Figure \ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters and with
two population size changes and subsequently estimated the posterior
distribution using a uniform prior for each slice independently.

The observation that we can associate a likelihood with any valid tree sequences
urges the need for some caveats. By treating each edge independently we can
naturally deal with the various degrees of completeness to which inheritance
is documented by the different ARG inference methods: e.g.\ polytomies, absence
of locally unary coalescence nodes, unpreserved identify of internal nodes
across marginal trees, \dots However, these imprecisions do violate the implicit
assumptions of the likelihood computation.
\comment{Do we want these remarks in here, or as part of the Discussion?}
For example, the described algorithm relies heavily on the
presence of locally unary nodes in the tree sequence to correctly identify the
full extent of all lineages going back through time and therefore recombination
events. Without unary nodes though, our algorithm will detect a recombination
event along both coalescing lineages in case of a (sub)tree height changes along
the ARG as both edges would switch parent nodes. This can be mitigated by limiting
the number of 'inferrable' recombination events per coalescence event to 1
(see fig \ref{sup:fig:rec-correction}).

>>>>

The contributions to the likelihood from coalescent and recombination events
is straightforward, since each possible coalescence is associated with rate $\lambda$,
and each recombination with rate $r$.
Under the SMC, we only ever have two lineages coalescing at a time;
however, inferred ARGs may have more-than-binary mergers.
We treat these as if there were zero-length edges
between a sequence of mergers;
the order of these unresolved mergers does not affect the likelihood.
(Under the SMC the likelihood of a zero-length edge is zero;
but this is not special: since edge lengths are continuously distributed,
the likelihood of any particular value is also zero.)
Let $\deg_o(n) = \#\{e_i \in E : p_i=n\}$ be the out-degree of a node $n$,
i.e., the number of edges having $n$ as a parent.
Since mergers are binary in the SMC,
this must have been the result of $\deg(n) - 1$ coalescences,
and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood
for nodes with any children.
Similarly,
let $\deg_i(n) = \#\{e_i \in E : c_i=n\}$ be the in-degree of a node $n$,
i.e., the number of edges having $n$ as a child.
Each additional edge that is ancestral to a given node
implies one additional ancestral recombination event,
and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood,
if the node has any parents.
Put together, these two contribute
\begin{equation}\label{eq:coal}
    \prod_{n \in N} \lambda^{\deg_o(n)-1} r^{\deg_i(n)-1}  
    =
    r^{|E|-|N|+n_r} 
    \lambda^{|E|-|N|+n_s} 
\end{equation}
to the likelihood,
where $n_r$ is the number of nodes with in-degree 0 (roots),
and $n_s$ is the number of nodes with out-degree 0 (usually, samples).
(The simple form follows because each edge contributes exactly one
to some node's out-degree and some node's in-degree.)

@@

The likelihood of ARG defined by the set of edges $E$ and nodes $N$
given parameters $\theta$ then is
\begin{equation}\label{eq:full-lik}
    \mathcal{L}(\mathcal{G}|\theta)
    =
    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(c, p) \in E} A_e(\theta) B_e(\theta) .
\end{equation}
Although the first two factors are easily expressed in terms of numbers of edges and nodes,
we will see below
that in practice it is convenient to absorb these terms into per-edge contributions.

@@

Next, we describe how to efficiently compute the likelihood under the SMC
from an ARG stored using the \emph{tree sequence} encoding. %  (see fig. \ref{fig:smc-unary}).
The tree sequence format encodes relationships using \emph{edges}
just as described above,
along with additional structure that
allows us to sequentially recover the local trees
along the genome very efficiently, and to reason about
the differences between those trees \citep{kelleher_efficient_2016, ralph_efficiently_2020}.
The order of the edge deletions and insertions required to move from one local tree $T_1$ to
the next, $T_2$ imposes the required strict total ordering on edges
(see definition \ref{def:coal})
that is needed to efficiently maintain the state of $I_e(t)$ for each focal edge $e$.
Indeed, following the removal of edges that do not persist in $T_2$,
we simply declare that each subsequently added edge
can only coalesce with the already-visited edges 
that make up the (partially) reconstructed local tree $T_2$.

\comment{This is \emph{very} nice!}
This allows us to compute a likelihood in a single pass over all edges.
More concretely, the likelihood is computed by maintaining a single vector $I$ that tracks
the number of extant lineages in each internode interval (see fig \ref{fig:algo}).
When moving from $T_1$ to $T_2$, the vector $I$ is first decremented along the internode intervals
spanned by the time of the child and the parent of all edges that do not persist in $T_2$.
For each subsequently inserted edge, we first compute the likelihood contribution
of that edge by computing the integral
in equation \eqref{eq:full-lik} using $I$, and then update the counts in $I$.
\comment{At first I thought we didn't need this next sentence now, but maybe
assigning-$\lambda$-and-$r$ to each edge makes it easier to recompute the likelihood
for slices?}
We further keep track of
the last parent a child node was connected to to detect recombination events.

Note that we can accommodate for stepwise changes in the coalescence rate (or recombination rate)
through time by intersecting the internode intervals with the rate change intervals.
Rate changes along the genome can be incorporated similarly (see \ref{sec:slicing}).

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG taken in
time and/or along the genome. This tool therefore strengthens
the inherent ability of ARGs to provide us with temporal
resolution and context to a dataset of genetic variation.
It also opens up the opportunity for more efficient optimization
strategies during inference given that each slice of the ARG can be
processed independently.

From a data structure perspective, we can compute a likelihood
under the SMC for any valid tree sequence.
The \tskit library provides a wide range
of subsetting operations on tree sequences allowing us to slice
the ARG in time and/or space, as well as to reduce the number of
tracked samples. Any such subset of the tree sequence is itself
a valid tree sequence for which we can again compute a likelihood.
\comment{This next remark is totally not obvious, and a really nice observation;
maybe it needs a bit more explanation?}
\comment{Oh, I think it's not \emph{quite} true: if we slide at position $x$,
then any recombinations that happened at $x$ are now invisible, right?
So maybe you need to slice at non-integer positions, or do some bookeeping
about the number of edges with starts exactly at $x$?}
Remarkably, if we slice an ARG into disjoint pieces,
the likelihood of the full ARG
is the product of the likelihoods of each of the pieces.

Figure \ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters with
two effective population size changes, and subsequently computed the posterior
distribution on the three effective population sizes
using a uniform prior on each slice independently.
The result is reasonably accurate,
despite only using 10kb of sequence.

Although we can associate a likelihood with any valid tree sequences,
there are some caveats. By treating each edge independently we can
naturally deal with the various degrees of completeness to which inheritance
is documented by the different ARG inference methods: e.g., polytomies, absence
of locally unary coalescence nodes, unpreserved identify of internal nodes
across marginal trees, etcetera.
However, many of these imprecisions violate the implicit
assumptions of the likelihood computation.
\comment{Do we want these remarks in here, or as part of the Discussion?}
For example, the described algorithm assumes the
presence of ``locally unary'' nodes to correctly identify the
full extent of all lineages going back through time and therefore recombination events.
Without unary nodes though, our algorithm will detect a recombination
event along both coalescing lineages in case of a (sub)tree height changes along
the ARG as both edges would switch parent nodes. This can be mitigated by limiting
the number of 'inferrable' recombination events per coalescence event to 1
(see fig \ref{sup:fig:rec-correction}).
\comment{I'm not editing this in since I'm not sure I've got it right:
I think that as written, our algorithm would count an extra recombination
and and extra coalescence for any edge that got split up by removing the unary bits.
This could be corrected for in the algorithm, but it would require extra bookkeeping,
and make the likelihood not so easily a product over edges. (?)}
